Multilevel computation of glacier geometry \\ from Stokes dynamics

### 2021-06-01 paper/msg.tex
A bit more MCD theory.

We will assume that $\Phi'(s)[t]$ is in the dual space of $W^{1,\qq}(\Omega)$ so that \eqref{eq:sigpfunctionalderiv} is finite.  Observe that $F'$ is straightforward to compute from $\Phi'$, but further comments on \eqref{eq:idoderiv} are forthcoming.  In fact, $F'(s)[t,r]$ and $\Phi'(s)[t]$ are one-sided directional derivatives in the sense that $s+\eps t$ must be admissible, i.e.~$s+\eps t\ge b$, for all sufficiently-small $\eps>0$, which requires $t\ge 0$ on the (active) set where $s=b$.  However, to simplify our considerations we define an $s$-independent set
\begin{equation}
\mathcal{D}_+ = \{t \in W_0^{1,\qq}(\Omega) \,:\, t(x,y) \ge 0\}. \label{eq:infdefectset}
\end{equation}
It follows that $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{D}_+$ and $F'(s)[t,r]$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{D}_+ \times W^{1,\qq}(\Omega)$.  We will approximate the true derivative $\Phi'(s)[t]$ by the difference quotient in \eqref{eq:idoderiv}, though only in the situation where $t \in \mathcal{D}_+$ has small support.

@@

Denote an arbitrary vertex of $\mathcal{T}$ by $(x_j,y_j)$ and let $\psi_j \in \mathcal{V}^h$ be the corresponding $P_1$ hat function, with $\psi_j(x_k,y_k)=\delta_{jk}$ \cite{Elmanetal2014}.  Let $\mathcal{D}_+^h = \{t^h \in \mathcal{V}^h \,:\, t^h \ge 0 \text{ and } t^h|_{\partial\Omega} = 0\}$.  Let $i=0,\dots,m-1$ index the interior vertices of $\mathcal{T}$, so $(x_i,y_i) \notin \partial\Omega$.  Note that $\psi_i \in \mathcal{D}_+^h$, and $s^h + c \psi_i \in \mathcal{K}^h$ if and only if $c\ge \beta_i^{s^h} = b^h(x_i,y_i) - s^h(x_i,y_i)$.  We call $\beta_i^{s^h}$ the pointwise defect obstacle \cite{GraeserKornhuber2009} for $s^h$.

@@

In this section we propose a multilevel scheme for solving the discrete SIGP weak form \eqref{eq:fesigpweakform} using iterated V-cycles of a new method descended from the multilevel constraint decomposition (MCD) method of \cite{Tai2003} (see also \cite{GraeserKornhuber2009}).  Our nonlinear variant of MCD, denoted MCDN, uses a full approximation scheme (FAS) multigrid \cite{Trottenbergetal2001} approach which transfers both the current residual and an approximation of the solution down to coarser levels.

We need notation for multiple mesh levels.  Suppose $\mathcal{T}^0$ is a fixed triangulation of $\Omega$, the coarse level.  Suppose $\{\mathcal{T}^j\}_{j=0}^J$ are standard uniform refinements of $\mathcal{T}^0$ by edge bisection so that each $T \in \mathcal{T}^j$ becomes four similar triangles in $\mathcal{T}^{j+1}$ \cite{Braess2007}.  (If $d=1$ this becomes simple halving of each interval.)  We seek a solution of \eqref{eq:fesigpweakform} on the fine level $\mathcal{T}^J$.  Let $\mathcal{V}^j$ be the $P_1$ FE space on $\mathcal{T}^j$ and $m_j$ be the number of nodes in, and dimension of, $\mathcal{V}^j$.

Suppose $b^J \in \mathcal{V}^J$ denotes the fine-level bed topography.  On this fine level the admissible surface elevations form a closed and convex set $\mathcal{L}^J = \{r^J \ge b^J, r^J|_{\partial\Omega} = b^J|_{\partial\Omega}\} \subset \mathcal{V}^J$.  However, instead of defining such admissible sets on each level, presumably via coarse interpolants of $b^J$, we follow \cite{GraeserKornhuber2009} in creating a multilevel decomposition starting from the defect obstacle for the current fine-level iterate $s^J$.  That is, suppose FIXME

FIXME Figure \ref{fig:decompclassical} shows such a decomposition

% figure generated in mg-glaciers/py by using decomposition_plain() in visualize.py and then:
% $ ./obstacle.py -J 5 -jcoarse 1 -irtol 1.0e-5 -monitor -diagnostics -random -randommodes 15 -o classical.pdf
% to get decomp_classical.pdf
\begin{figure}[t]
\begin{center}
\includegraphics[width=0.7\textwidth]{fixfigs/decompclassical.pdf}
\end{center}
\caption{An example $d=1$ and $J=4$ MCD decomposition of the defect obstacle $\chi^J$ for an iterate $s^J$.}
\label{fig:decompclassical}
\end{figure}

Before describing the MCDN approach we should first be clear about when an iteration has converged.  For a discrete obstacle $\varphi^j \in \mathcal{V}^j$, an admissible current iterate $s^j\in \{r^j \ge \varphi^j\} \subset \mathcal{V}^j$, and a residual vector $F^j \in (\mathcal{V}^j)'$ let
\begin{equation}
\vertiii{F^j}_{(s^j,\varphi^j)} = \left(\sum_{s_i > \varphi_i} |F[\psi_i]|^2 + \sum_{s_i = \varphi_i} |\min\{F[\psi_i],0\}|^2\right)^{1/2} \label{eq:cpnorm}
\end{equation}
be the ``CP residual norm'' associated to the finite-dimensional complementarity problem (CP) $s^j \ge \varphi^j$, $F(s^j) \ge 0$, and $(s^j-\varphi^j) F(s^j) = 0$.  Note $s^j \in \mathcal{V}^j$ solves this CP if and only if $s^j$ is admissible and $\vertiii{F(s^j)}_{(s^j,\varphi^j)}=0$.  As shown in Pseudocode \ref{pc:mcdn-solver} we will iterate MCDN V-cycles until the CP residual norm is small according to either an absolute or a relative tolerance.

@@

FIXME nonlinear MCD V-cycle Pseudocode \ref{pc:mcdn-vcycle}

\begin{pcode}[ht]
\begin{pseudo*}
\pr{mcdn-vcycle}(J,w^J,N,\ell^J,\chi^J,\id{down}=1,\id{coarse}=1,\id{up}=0)\text{:} \\+
    $g^J = w^J$ \\
    for $j=J$ downto $j=1$ \\+
      $\chi^{j-1} = \mR \chi^j$ \\
      $\phi^j = \chi^j - P\chi^{j-1}$ \\
      $y^j = 0$ \\
      $\text{\pr{smoother}}^{\text{\id{down}}}(j,g^j,y^j,N^j,\ell^j,\phi^j)$ \qquad \ct{smoothing in $\mathcal{K}^j$} \\
      $F = N^j(g^j+y^j) - \ell^j$ \\
      $g^{j-1} = \iR(g^j + y^j)$ \\
      $\ell^{j-1} = N^{j-1}(g^{j-1}) - R F$ \\-
    $y^0 = 0$ \\
    $\text{\pr{smoother}}^{\text{\id{coarse}}}(0,g^0,y^0,N^0,\ell^0,\chi^0)$ \\
    $z^0 = y^0$ \\
    for $j=1$ to $j=J$ \\+
      $z^j = P z^{j-1} + y^{j}$ \\
      $\text{\pr{smoother}}^{\text{\id{up}}}(j,g^j,z^j,N^j,\ell^j,\chi^j)$ \qquad \ct{smoothing in $\mathcal{D}^j$} \\-
    return $z^J$
\end{pseudo*}
\caption{FIXME Nonlinear MCD V-cycle.}
\label{pc:mcdn-vcycle}
\end{pcode}

@@

\renewcommand{\arraystretch}{1.1}
\begin{longtable}{l|l|l}
\toprule
\textbf{Acronym} {\Large$\strut$} & \textbf{Definition} & \textbf{Reference} \\ \hline
CMB & climatic mass balance & Section \ref{sec:stokesgeometry} \\
FAS & full approximation scheme & Section \ref{sec:mcdstokes} \\
FE & finite element & Section \ref{sec:fe} \\
GS & Gauss-Seidel & Section \ref{sec:smoothers} \\
IDO & ice dynamics operator & Section \ref{sec:weakido}, equation \eqref{eq:ido} \\
IIGP & implicit ice geometry problem & Section \ref{sec:evolution} \\
MCD & multilevel constraint decomposition & Section \ref{sec:mcdstokes} \\
MCDN & multilevel constraint decomposition (nonlinear) & Section \ref{sec:mcdstokes}, Pseudocode \ref{pc:mcdn-vcycle} \\
NCP & nonlinear complementarity problem & Section \ref{sec:stokesgeometry} \\
PNGS & projected, nonlinear Gauss-Seidel (smoother) & Section \ref{sec:smoothers}, Pseudocode \ref{pc:pngsslow} \\
PNJ & projected, nonlinear Jacobi (smoother) & Section \ref{sec:smoothers}, Pseudocode \ref{pc:pnj} \\
SIA & shallow ice approximation & Section \ref{sec:intro} \\
SIGP & steady ice geometry problem & Section \ref{sec:stokesgeometry}, equation \eqref{eq:strongform} \\
SKE & surface kinematical equation & Section \ref{sec:stokesgeometry}, equation \eqref{eq:ske} \\
VI & variational inequality & Section \ref{sec:weakido} \\
WU & work units & Section \ref{sec:mcdstokes} \\ % final \\ required
\bottomrule
\caption{Glossary of acronyms used in this paper.}
\label{tab:acronyms}
\end{longtable}

>>>>

We will assume that $\Phi'(s)[t]$ is in the dual space of $W^{1,\qq}(\Omega)$ so that \eqref{eq:sigpfunctionalderiv} is finite.  Observe that $F'$ is straightforward to compute from $\Phi'$, but further comments on \eqref{eq:idoderiv} are forthcoming.  In fact, $F'(s)[t,r]$ and $\Phi'(s)[t]$ are one-sided directional derivatives in the sense that $s+\eps t$ must be admissible, i.e.~$s+\eps t\ge b$, for all sufficiently-small $\eps>0$, which requires $t\ge 0$ on the (active) set where $s=b$.  However, to simplify our considerations we define an $s$-independent set
\begin{equation}
\mathcal{W}_+ = \{t \in W_0^{1,\qq}(\Omega) \,:\, t(x,y) \ge 0\}. \label{eq:infdefectset}
\end{equation}
It follows that $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{W}_+$ and $F'(s)[t,r]$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{W}_+ \times W^{1,\qq}(\Omega)$.  We will approximate the true derivative $\Phi'(s)[t]$ by the difference quotient in \eqref{eq:idoderiv}, though only in the situation where $t \in \mathcal{W}_+$ has small support.

@@

Denote an arbitrary vertex of $\mathcal{T}$ by $(x_j,y_j)$ and let $\psi_j \in \mathcal{V}^h$ be the corresponding $P_1$ hat function, with $\psi_j(x_k,y_k)=\delta_{jk}$ \cite{Elmanetal2014}.  Let $\mathcal{W}_+^h = \{t^h \in \mathcal{V}^h \,:\, t^h \ge 0 \text{ and } t^h|_{\partial\Omega} = 0\}$.  Let $i=0,\dots,m-1$ index the interior vertices of $\mathcal{T}$, so $(x_i,y_i) \notin \partial\Omega$.  Note that $\psi_i \in \mathcal{W}_+^h$, and $s^h + c \psi_i \in \mathcal{K}^h$ if and only if $c\ge \beta_i^{s^h} = b^h(x_i,y_i) - s^h(x_i,y_i)$.  We call $\beta_i^{s^h}$ the pointwise defect obstacle \cite{GraeserKornhuber2009} for $s^h$.

@@

In this section we propose a new multilevel scheme for solving the discrete SIGP weak form \eqref{eq:fesigpweakform} using iterated V-cycles.  The method descends from the multilevel constraint decomposition (MCD) method of \cite{Tai2003} (see also \cite{GraeserKornhuber2009}).  Our nonlinear variant, denoted MCDN, uses a full approximation scheme (FAS) multigrid \cite{Trottenbergetal2001} approach which transfers both the current residual and an approximation of the solution down to coarser levels.

Regarding notation for multiple mesh levels, suppose $\mathcal{T}^0$ is a fixed triangulation of $\Omega$, and call it the coarse level.  For $J\ge 0$, suppose $\{\mathcal{T}^j\}_{j=1}^J$ are standard uniform refinements of $\mathcal{T}^0$ by edge bisection so that each $T \in \mathcal{T}^j$ becomes four similar triangles in $\mathcal{T}^{j+1}$ \cite{Braess2007}.  (We assume simple halving of each interval when $d=1$.)  Let $\mathcal{V}^j$ be the $P_1$ FE space on $\mathcal{T}^j$ and let $\mathcal{V}_0^j \subset \mathcal{V}^j$ be the $P_1$ functions which are zero on $\partial\Omega$.  If $m_j$ is the number of interior nodes in $\mathcal{T}$ then $\dim(\mathcal{V}_0^j)=m_j$.

We seek an admissible solution to \eqref{eq:fesigpweakform} on the fine level $\mathcal{T}^J$.  Suppose $b^J \in \mathcal{V}^J$ denotes the fine-level bed topography.  On this fine level the admissible surface elevations form a closed and convex set
\begin{equation}
\mathcal{K}^J = \{r^J \in \mathcal{V}^J \,:\, r^J \ge b^J, r^J|_{\partial\Omega} = b^J|_{\partial\Omega}\}.  \label{eq:singleadmissible}
\end{equation}
However, we follow \cite{GraeserKornhuber2009} in creating a multilevel decomposition of problem \eqref{eq:fesigpweakform} based on a ``defect obstacle'' formulation, and not direct use of the set $\mathcal{K}^J$.  That is, suppose $s^J \in \mathcal{K}^J$ is an admissible current iterate and let
\begin{equation}
\chi^J = b^J - s^J \in \mathcal{V}_0^J \label{eq:finedefectobstacle}
\end{equation}
be the associated defect obstacle, a nonpositive function on $\Omega$.  Note that $z^J \in \mathcal{V}_0^J$ is an admissible perturbation of $s^J$, i.e.~$s^J+z^J \in \mathcal{K}^J$, if and only if $z^J \ge \chi^J$.

Also following \cite{GraeserKornhuber2009}, we define the monotone restriction operator $\mR : \mathcal{V}_0^j \to \mathcal{V}_0^{j-1}$.  Suppose $z^j \in \mathcal{V}_0^j$ is expanded in interior-node hat functions $\psi_i^j$, and denote the level-$j$ nodes as $(x_i^j,y_i^j)$.  If $z^j = \sum_{i=0}^{m_j-1} z^j[i] \psi_i^j$ for coefficients $z^j[i]\in \RR$ then
\begin{equation}
\mR z^j = \sum_{\ell=0}^{m_{j-1}-1} \max\left\{z^j[i] \,:\,\psi_\ell^{j-1}(x_i^j,y_i^j) \right\}\,\psi_\ell^{j-1}.  \label{eq:monotonerestriction}
\end{equation}
In words, the coefficients for $\mR z^j \in \mathcal{V}_0^j$ are computed by maximizing the finer-level nodal values over the (open) supports of each each coarser-level hat function.

Now we can recursively define a defect obstacle on each level:
\begin{equation}
\chi^{j-1} = \mR \chi^j  \label{eq:recursivedefectobstacle}
\end{equation}
for $j=1,\dots,J$, as shown for example in Figure \ref{fig:decompclassical}.  We also use the (nonpositive) differences of the defect obstacles,
\begin{equation}
\phi^j = \chi^j - \chi^{j-1}.  \label{eq:downobstacles}
\end{equation}
In terms of these functions, for $j=0,\dots,J$ we define down-constraint sets
\begin{equation}
\mathcal{D}^j = \left\{z^j \in \mathcal{V}_0^j \,:\, z^j \ge \chi^j\right\}.  \label{eq:downconstraintset}
\end{equation}
and up-constraint sets
\begin{equation}
\mathcal{U}^j = \left\{z^j \in \mathcal{V}_0^j \,:\, z^j \ge \phi^j\right\}.  \label{eq:upconstraintset}
\end{equation}
This terminology of down- and up-constraint sets is explained by noting that the smoother will solve one-dimensional VIs on each level, as described in the previous section, where the admissible sets are different as one goes down and then up in a V-cycle.

% figure generated in mg-glaciers/py by using decomposition_plain() in visualize.py and then:
% $ ./obstacle.py -J 5 -jcoarse 1 -irtol 1.0e-5 -monitor -diagnostics -random -randommodes 15 -o classical.pdf
% to get decomp_classical.pdf
\begin{figure}[t]
\begin{center}
\includegraphics[width=0.7\textwidth]{fixfigs/decompclassical.pdf}
\end{center}
\caption{An example ($d=1$, $J=4$) of an MCD decomposition of the defect obstacle $\chi^J$.  Defect obstacles $\chi^j$ are generated using the monotone restriction operator $\mR$ as in \eqref{eq:recursivedefectobstacle}.}
\label{fig:decompclassical}
\end{figure}

FIXME DEBUG THIS PSEUDOCODE nonlinear MCD V-cycle Pseudocode \ref{pc:mcdn-vcycle}

\begin{pcode}[ht]
\begin{pseudo*}
\pr{mcdn-vcycle}(J,s^J,b^J,\ell^J,\id{down}=0,\id{coarse}=1,\id{up}=1)\text{:} \\+
    $\chi^J = b^J - s^J$ \\
    $g^J = s^J$ \\
    for $j=J$ downto $j=1$ \\+
      $\chi^{j-1} = \mR \chi^j$ \\
      $\phi^j = \chi^j - \chi^{j-1}$ \\
      $y^j = 0$ \\
      $\text{\pr{smoother}}^{\text{\id{down}}}(y^j,g^j,\ell^j,\phi^j)$ \qquad \ct{smoothing in $\mathcal{D}^j$} \\
      $f = F^j(g^j+y^j) - \ell^j$ \\
      $g^{j-1} = \iR(g^j + y^j)$ \\
      $\ell^{j-1} = F^{j-1}(g^{j-1}) - R f$ \\-
    $y^0 = 0$ \\
    $\text{\pr{smoother}}^{\text{\id{coarse}}}(y^0,g^0,\ell^0,\chi^0)$ \qquad \ct{smoothing in $\mathcal{U}^0$} \\
    $z^0 = y^0$ \\
    for $j=1$ to $j=J$ \\+
      $z^j = P z^{j-1} + y^{j}$ \\
      $\text{\pr{smoother}}^{\text{\id{up}}}(z^j,g^j,\ell^j,\chi^j)$ \qquad \ct{smoothing in $\mathcal{U}^j$} \\-
    return $z^J$
\end{pseudo*}
\caption{FIXME Nonlinear MCD V-cycle.}
\label{pc:mcdn-vcycle}
\end{pcode}

Finally, we need to be clear about when an iteration has converged.  For a discrete obstacle $\varphi^j \in \mathcal{V}^j$, an admissible current iterate $s^j\in \{r^j \ge \varphi^j\} \subset \mathcal{V}^j$, and a residual vector $F^j \in (\mathcal{V}^j)'$ let
\begin{equation}
\vertiii{F^j}_{(s^j,\varphi^j)} = \left(\sum_{s_i > \varphi_i} |F[\psi_i]|^2 + \sum_{s_i = \varphi_i} |\min\{F[\psi_i],0\}|^2\right)^{1/2} \label{eq:cpnorm}
\end{equation}
be the ``CP residual norm'' associated to the finite-dimensional complementarity problem (CP) $s^j \ge \varphi^j$, $F(s^j) \ge 0$, and $(s^j-\varphi^j) F(s^j) = 0$.  Note $s^j \in \mathcal{V}^j$ solves this CP if and only if $s^j$ is admissible and $\vertiii{F(s^j)}_{(s^j,\varphi^j)}=0$.  As shown in Pseudocode \ref{pc:mcdn-solver} we will iterate MCDN V-cycles until the CP residual norm is small according to either an absolute or a relative tolerance.

@@

@@

\renewcommand{\arraystretch}{1.1}
\begin{longtable}{l|l|l}
\caption{Glossary of acronyms used in this paper.}
\label{tab:acronyms} \\ % \\ REQUIRED HERE
\toprule
\textbf{Acronym} {\Large$\strut$} & \textbf{Definition} & \textbf{Reference} \\ \hline
CMB & climatic mass balance & Section \ref{sec:stokesgeometry} \\
FAS & full approximation scheme & Section \ref{sec:mcdstokes} \\
FE & finite element & Section \ref{sec:fe} \\
GS & Gauss-Seidel & Section \ref{sec:smoothers} \\
IDO & ice dynamics operator & Section \ref{sec:weakido}, equation \eqref{eq:ido} \\
IIGP & implicit ice geometry problem & Section \ref{sec:evolution} \\
MCD & multilevel constraint decomposition & Section \ref{sec:mcdstokes} \\
MCDN & multilevel constraint decomposition (nonlinear) & Section \ref{sec:mcdstokes}, Pseudocode \ref{pc:mcdn-vcycle} \\
NCP & nonlinear complementarity problem & Section \ref{sec:stokesgeometry} \\
PNGS & projected, nonlinear Gauss-Seidel (smoother) & Section \ref{sec:smoothers}, Pseudocode \ref{pc:pngsslow} \\
PNJ & projected, nonlinear Jacobi (smoother) & Section \ref{sec:smoothers}, Pseudocode \ref{pc:pnj} \\
SIA & shallow ice approximation & Section \ref{sec:intro} \\
SIGP & steady ice geometry problem & Section \ref{sec:stokesgeometry}, equation \eqref{eq:strongform} \\
SKE & surface kinematical equation & Section \ref{sec:stokesgeometry}, equation \eqref{eq:ske} \\
VI & variational inequality & Section \ref{sec:weakido} \\
WU & work units & Section \ref{sec:mcdstokes} \\ % final \\ required
\bottomrule
\end{longtable}
